Statistical methods for relative quantification of post-translational modifications in global proteomics experiments

### 2021-10-07 supplementary/ptm_sm.tex
Updated main per Olgas comments, added sample size analysis

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Computer simulation 1 - label free experiment}

@@

\subsubsection{Computer simulation 2 - missing values and low replicates}

@@

\clearpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{SpikeIn benchmark - KGG enriched}
\label{sec:benchmark}

@@

\subsection{Human-1mix-TMT - ubiquitination}
\label{sec:ipah}

@@

\clearpage
\subsection{Mouse-2mix-TMT - phosphorylation}
\label{sec:shigella}

@@

\subsection{Human label free quantification - no global profiling run}
\label{sec:usp30}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{sim_new/USP30_profile_plot}
\caption{Comparing the global profiling of protein $P52209$ with the modification of the protein at site $K059$. The modification appears generally unchanged between all conditions, whereas the global profiling run shows the CCCP and Combo conditions have a higher relative abundance compared to the Control and USP30\_OE. This indicates that the modification actually had a major effect when comparing CCCP and Combo to Control and USP30\_OE. This would have been missed without adjusting for global protein changes.}
\label{fig:USP30_profile_plot}
\end{figure}
\clearpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sample size calculation and power analysis}
\todo{Devon: Update this section per Olga's commentsv}

The proposed approach allows us to calculate the sample size needed to achieve a desired statistical power as described in Section \ref{sec:design}. As described in Section \ref{sec:test} the proposed method adjusts for the underlying protein abundance in the PTM significance analysis, which corrects the confounding factor with a cost of increased variation. When the variation is increased, the experiment requires a larger number of replicates to reach the same power. Thus the statistical power is dependent on the variance from both the modified peptide and unmodifed protein as well as the number of replicates in both the modified and unmodified runs.

We compared the statistical power in experiments with different numbers of replicates, variance, and fold change for both the modified and unmodified runs. The results can be seen in \sfigref{fig:power_sd_combo}

\begin{figure}[h!]
\centering
\includegraphics[width=.9\textwidth]{sim_new/power_analysis_sd_combined}
\caption{}
\label{fig:power_sd_combo}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Dataset 1 : Computer simulation 1 - label free experiment}

@@

\subsubsection{Dataset 2 : Computer simulation 2 - missing values and low replicates}

@@

\clearpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Dataset 3 : SpikeIn benchmark - label-free - KGG enriched}
\label{sec:benchmark}

@@

\subsection{Dataset 4 : Human-1mix-TMT - ubiquitination}
\label{sec:ipah}

@@

\clearpage
\subsection{Dataset 5 : Mouse-2mix-TMT - phosphorylation}
\label{sec:shigella}

@@

\subsection{Dataset 6 : Human label free quantification - no global profiling run}
\label{sec:usp30}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{sim_new/USP30_profile_plot}
\caption{Comparing the global profiling of protein $P52209$ with the modification of the protein at site $K059$. The modification appears generally unchanged between all conditions, whereas the global profiling run shows the CCCP and Combo conditions have a higher relative abundance compared to the Control and USP30\_OE. This indicates that the modification actually had a major effect when comparing CCCP and Combo to Control and USP30\_OE. This would have been missed without adjusting for global protein changes.}
\label{fig:USP30_profile_plot}
\end{figure}
\clearpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sample size calculation and power analysis}

The proposed approach allows us to calculate the sample size needed to achieve a desired statistical power as described in Section \ref{sec:design}. The proposed method adjusts for the underlying protein abundance in the PTM significance analysis, which corrects the confounding factor with a cost of increased variation. When the variation is increased, the experiment requires a larger number of replicates to reach the same power. Thus the statistical power is dependent on the variance from both the modified peptide and unmodifed protein as well as the number of replicates in both runs.

We compared the statistical power in experiments with differing numbers of replicates, variance, and fold change for both the modified and unmodified runs. For variance we used the variance from the biological experiments to determine what values to set. In Dataset 4 and 5 the variance of the PTM was generally higher than the global protein. In contrast, in dataset 6 the variance of the PTM and Protein was generally the same. Because of this we will analyze the power of our experiments when the PTM variance is higher than the protein, and also when they are the same. For the variance numbers, in the case when the PTM and protein are the same we chose a variance of .15. When the PTM is higher than the protein we chose a PTM variance of .2 and a protein variance of .1. These numbers were derived from the biological experiments in this paper.

The results can be seen in \sfigref{fig:power_sd_combo}

\begin{figure}[h!]
\centering
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.9\textwidth]{sim_new/same_var_power}
	\caption{blah}
 \end{subfigure}\vspace{-5mm}
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.9\textwidth]{sim_new/high_ptm_var_power}
	\caption{blah}
 \end{subfigure}
 \caption{Power analysis of experiments with differing variances.}
\label{fig:power_sd_combo}
\end{figure}
